Bool literals are designated by the following syntax:
\begin{syntax}
bool-literal: one of
  `true' `false'
\end{syntax}

Signed and unsigned integer literals are designated by the following
syntax:
\begin{syntax}
integer-literal:
  digits
  0 'x' hexadecimal-digits
  0 'b' binary-digits

digits:
  digit
  digit digits

digit: one of
  0 1 2 3 4 5 6 7 8 9

hexadecimal-digits:
  hexadecimal-digit
  hexadecimal-digit hexadecimal-digits

hexadecimal-digits: one of
  0 1 2 3 4 5 6 7 8 9 A B C D E F a b c d e f

binary-digits:
  binary-digit
  binary-digit binary-digits

binary-digits: one of
  0 1
\end{syntax}
Suffixes, like those in C, are not necessary.  The type of an integer
literal is the first type of the following that can hold the value of
the digits: \chpl{int}, \chpl{int(64)}, \chpl{uint(64)}.  Explicit
conversions are necessary to change the type of the literal to another
integer size.

Real literals are designated by the following syntax:
\begin{syntax}
real-literal:
  digits[OPT] . digit digits[OPT] exponent-part[OPT]

exponent-part:
  `e' sign[OPT] digits

sign: one of
  + -
\end{syntax}
The type of a real literal is \chpl{real}.  Explicit conversions are
necessary to change the type of the literal to another real
size.

Note that real literals require that a digit follow the
decimal point.  This is necessary to avoid an ambiguity in
interpreting \chpl{2.e+2} that arises if a method called \chpl{e} is
defined on integers.

Imaginary literals are designated by the following syntax:
\begin{syntax}
imaginary-literal:
  real-literal `i'
  integer-literal `i'
\end{syntax}

A complex number is specified by adding or subtracting an imaginary
literal with a real literal.  Alternatively, a 2-tuple literal of
expressions of integer or real type can be cast to a complex.  These
expressions can be literals, but do not need to be.  To create a
complex literal or parameter, they must be literals or parameters.

\begin{example}
The following codes represent the same complex literal:
\begin{center}
\chpl{2.0i}, \hspace{1pc} \chpl{0.0+2.0i}, \hspace{1pc}
\chpl{(0.0,2.0):complex}.
\end{center}
\end{example}

String literals are designated by the following syntax:
\begin{syntax}
string-literal:
  " characters "
  ' characters '

characters:
  character
  character characters

character:
  any-character
\end{syntax}

\begin{implementation}
Strings are currently restricted to ASCII characters.  In a future
version of Chapel, strings will be defined over alphabets to allow for
more exotic characters.
\end{implementation}
