\documentclass[11pt]{article}


%===> Packages ===>
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%<=== Packages <===


%===> New commands ===>
\newcommand{\dx}{\Delta x}
\newcommand{\dt}{\Delta t}
\newcommand{\bigoh}{\mathcal{O}}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\bslash}{\char`\\}
\newcommand{\uscore}{\char`\_}
%<=== New commands <===


%===> Formatting ===>
\setlength{\parskip}{11pt}
\setlength{\parindent}{0pt}
%<=== Formatting <===


\begin{document}


\section*{Single grid problems}

\subsection*{Miscellaneous}

\texttt{tuplify} routine

\texttt{intersectDomains} routine

\texttt{GridArray} class

\texttt{IsolatedArray} class



\subsection*{BaseGrid class}

\subsubsection*{Fields}

\texttt{x\uscore{}low, x\uscore{}high}: coordinate bounds

\texttt{i\uscore{}low, i\uscore{}high}: index bounds

\texttt{dx}: mesh width in each dimension

\texttt{cells}: domain of interior cells

\texttt{ext\uscore{}cells}: \texttt{cells}, plus ghost cells

\texttt{low\uscore{}ghost\uscore{}cells}: Tuple of domains, each specifying the ghost cells at the low end of a dimension.

\texttt{high\uscore{}ghost\uscore{}cells}: Ditto, but for the high end.



\subsubsection*{Methods}

\texttt{xValue}: Returns coordinates of an index

\texttt{initializeSolution}: Initializes a \texttt{ScalarGridSolution}.

\texttt{clawOutput}: Writes out a \texttt{ScalarGridSolution} in Clawpack format.  \textit{Should this act on an array instead?}



\subsubsection*{Subclasses}

\texttt{GhostData}: Class for holding data on \textbf{only} the ghost cells.  \textit{Should this be a subclass?}




\subsection*{\texttt{ScalarGridSolution} class}

\subsubsection*{Fields}

\texttt{grid}: \texttt{BaseGrid} on which the solution is defined.

\texttt{space\uscore{}data}: \texttt{[1..2] [grid.ext\uscore{}cells] real}
Holds two time levels of data on the parent grid

\texttt{time}: \texttt{[1..2] real}  Times at which \texttt{space\uscore{}data} is specified.



\end{document}

GridBC class:
=============
Fields:
-------
    grid: Grid on which the BCs are defined.
    low_boundary_faces: Tuple of domains, each specifying the faces at the low end of a dimension.
    high_boundary_faces: Ditto, but for the high end.

Methods:
--------
    ghostFill: Fills ghost cells at a given time.
    homogeneousGhostFill: The homogeneous method corresponding to ghostFill.  Necessary for
        implicit solvers.


\end{document}


\section*{Intent}
\begin{itemize}
\item Demonstrate Chapel's ability to simplify development of AMR codes.
\item Write generic-dimensional code whenever possible --- drastically simplifies code base.
\item (Personal goal) Draft an AMR system where single-grid or single-level problems can be drafted without using unnecessary framework.  Should allow step-wise development of AMR solvers.
\end{itemize}

\section*{Validation}
\begin{itemize}
\item Chombo's \texttt{IntVec}s and \texttt{IntVecSet}s
\end{itemize}

\section*{Chapel features}


\end{document}
