\sekshun{Arrays}
\label{Arrays}
\index{arrays}

An \emph{array} is a map from a domain's indices to a collection of
variables.

Arrays abstract mappings from sets of values to variables.  This key
use of data structures coupled with the generic syntactic support for
array usage increases software reusability.  By separating the sets of
values into their own abstraction, \emph{i.e.}, domains, distributions
can be associated with sets rather than variables.  This enables the
orthogonality of data distributions.  Distributions are discussed
in~\rsec{Domain_Maps}.

\subsection{Arrays}
%\label{Arrays}
\index{arrays}

Arrays associate variables or elements with the sets of indices in a
domain.  Arrays must be declared over domains and have a specified
element type.

\subsubsection{Array Types}
\label{Array_Types}
\index{arrays!types}

The type of an array is parameterized by the type of the domain that
it is declared over and the element type of the array.  Array types
are given by the following syntax:
\begin{syntax}
array-type:
  [ domain-expression ] type-specifier

domain-expression:
  domain-literal
  expression
\end{syntax}
The \sntx{domain-expression} must specify a domain that the array can
be declared over.  This can be a domain literal.  If it is a domain
literal, the square brackets around the domain literal can be omitted.

\begin{example}
In the code
\begin{chapel}
var A: [D] real;
\end{chapel}
\chpl{A} is declared to be an array over domain \chpl{D} with
elements of type \chpl{real}.
\end{example}

An array's element type can be referred to using the member symbol
\chpl{eltType}.

\begin{example}
In the following example, \chpl{x} is declared to be of type
\chpl{real} since that is the element type of array \chpl{A}.
\begin{chapel}
var A: [D] real;
var x: A.eltType;
\end{chapel}
\end{example}

\subsubsection{Array Indexing}
\label{Array_Indexing}
\index{arrays!indexing}

Arrays can be indexed by indices in the domain they are declared over.
The indexing results in an access of the element that is mapped by
this index.

\begin{example}
If \chpl{A} is an array with element type \chpl{real} declared over a
one-dimensional arithmetic domain \chpl{[1..n]}, then the first
element in \chpl{A} can be accessed via the expression \chpl{A(1)} and
set to zero via the assignment \chpl{A(1) = 0.0}.
\end{example}

Indexing into an array with a domain is call array slicing and is
discussed in the next section.

Arithmetic arrays also support indexing over the components of their
indices for multidimensional arithmetic domains (where the indices are
tuples), as described in~\rsec{Arithmetic_Array_Indexing}.

One-dimensional arrays may be indexed by 1-tuples of their index
types.
\begin{example}
The code
\begin{chapel}
def f(A: [], is...)
  return A(is);
\end{chapel}
defines a function that takes an array as the first argument and a
variable-length argument list.  It then indexes into the array using
the tuple that captures the actual arguments.  This function works
even for one-dimensional arrays because one-dimensional arrays can be
indexed into by 1-tuples.
\end{example}

\subsubsection{Array Slicing}
\label{Array_Slicing}
\index{arrays!slicing}

An array can be indexed by a domain that has the same type as the
domain which the array was declared over.  Indexing in this manner has
the effect of array slicing.  The result is a new array declared over
the indexing domain where the elements in the array alias the elements
in the array being indexed.

\begin{example}
Given the definitions
\begin{chapel}
var OuterD: domain(2) = [0..n+1, 0..n+1];
var InnerD: domain(2) = [1..n, 1..n];
var A, B: [OuterD] real;
\end{chapel}
the assignment given by
\begin{chapel}
A(InnerD) = B(InnerD);
\end{chapel}
assigns the elements in the interior of \chpl{B} to the elements in
the interior of \chpl{A}.
\end{example}

Arithmetic arrays also support slicing by indexing into them with
ranges or tuples of ranges as described
in~\rsec{Arithmetic_Array_Slicing}.

\subsubsection{Array Assignment}
\label{Array_Assignment}
\index{arrays!assignment}

Array assignment is by value.  Arrays can be assigned arrays, ranges,
domains, iterators, or tuples.  If \chpl{A} is an lvalue of array type
and \chpl{B} is an expression of either array, range, or domain type,
or an iterator, then the assignment
\begin{chapel}
A = B;
\end{chapel}
is equivalent to
\begin{chapel}
forall (i,e) in (A.domain,B) do
  A(i) = e;
\end{chapel}
If the zipper iteration is illegal, then the assignment is illegal.
Notice that the assignment is implemented with the semantics of
a \chpl{forall} loop.

Arrays can be assigned tuples of values of their element type if the
tuple contains the same number of elements as the array.  For
multidimensional arrays, the tuple must be a nested tuple such that
the nesting depth is equal to the rank of the array and the shape of
this nested tuple must match the shape of the array.  The values are
assigned element-wise.

Arrays can also be assigned single values of their element type.  In
this case, each element in the array is assigned this value.
If \chpl{e} is an expression of the element type of the array or a
type that can be implicitly converted to the element type of the
array, then the assignment
\begin{chapel}
A = e;
\end{chapel}
is equivalent to
\begin{chapel}
forall i in A.domain do
  A(i) = e;
\end{chapel}

\subsubsection{Formal Arguments of Array Type}
\label{Formal_Arguments_of_Array_Type}
\index{arrays!as formal arguments}

Arrays are passed to functions by reference.  Formal arguments that
receive arrays are aliases of the actual arguments.  The ordinary rule
that disallows assignment to formal arguments of blank intent does not
apply to arrays.

When a formal argument has array type, the element type of the array
can be omitted and/or the domain of the array can be queried or
omitted.  In such cases, the argument is generic and is discussed
in~\rsec{Formal_Arguments_of_Generic_Array_Types}.

If a non-queried domain is specified in the array type of a formal
argument, the domain must match the domain of the actual argument.
This is verified at runtime.  There is an exception if the domain is
an arithmetic domain; it is described
in~\rsec{Formal_Arguments_of_Arithmetic_Array_Type}.

\subsubsection{Iteration over Arrays}
\label{Iteration_over_Arrays}

All arrays support iteration via forall and for loops over the
elements mapped to by the indices in the array's domain.

\subsubsection{Array Promotion of Scalar Functions}
\label{Array_Promotion_of_Scalar_Functions}
\index{arrays!promotion}

Array promotion of a scalar function is defined over the array type
and the element type of the array.  The domain of the returned array,
if an array is captured by the promotion, is the domain of the array
that promoted the function.  In the event of zipper promotion over
multiple arrays, the promoted function returns an array with a domain
that is equal to the domain of the first argument to the function that
enables promotion.  If the first argument is an iterator or a range,
the result is a one-based one-dimensional array.

\begin{example}
Whole array operations is a special case of array promotion of scalar
functions.  In the code
\begin{chapel}
A = B + C;
\end{chapel}
if \chpl{A}, \chpl{B}, and \chpl{C} are arrays, this code assigns each
element in \chpl{A} the element-wise sum of the elements in \chpl{B}
and \chpl{C}.
\end{example}

\subsubsection{Array Initialization}
\label{Array_Initialization}
\index{arrays!initialization}

By default, the elements in an array are initialized to the default
values associated with the element type of the array.  There is an
expectation that this default initialization can be overridden for
performance reasons by explicitly marking the array type or variable.

\subsubsection{Array Aliases}
\label{Array_Aliases}
\index{=>@\chpl{=>}}

Array slices alias the data in arrays rather than copying it.  Such
array aliases can be captured and optionally reindexed with the array
alias operator \chpl{=>}.  The syntax for capturing an alias to an
array requires a new variable declaration:
\begin{syntax}
array-alias-declaration:
  identifier reindexing-expression[OPT] => array-expression ;

reindexing-expression:
  [ domain-expression ]

array-expression:
  expression
\end{syntax}
The identifier is an alias to the array specified in
the \sntx{array-expression}.

The optional \sntx{reindexing-expression} allows the domain of the
array alias to be reindexed.  The shape of the domain in
the \sntx{reindexing-expression} must match the shape of the domain of
the \sntx{array-expression}.  Indexing via the alias is governed by
the new indices.

\begin{example}
In the code
\begin{chapel}
var A: [1..5, 1..5] int;
var AA: [0..2, 0..2] => A[2..4, 2..4];
\end{chapel}
an array alias \chpl{AA} is created to alias the interior of
array \chpl{A} given by the slice \chpl{A[2..4, 2..4]}.  The
reindexing expression changes the indices defined by the domain of the
alias to be zero-based in both dimensions.  Thus \chpl{AA(1,1)} is
equivalent to \chpl{A(3,3)}.
\end{example}

\subsection{Arithmetic Arrays}
\label{Arithmetic_Arrays}
\index{arrays!arithmetic}

An arithmetic domain is a rectilinear set of Cartesian indices.
Arithmetic domains are specified as a tuple of ranges enclosed in
square brackets.


\subsubsection{Arithmetic Array Indexing}
\label{Arithmetic_Array_Indexing}

In addition to being indexed by indices defined by their arithmetic
domains, arithmetic arrays can be indexed directly by values of the
dimensional index type where the number of values is equal to the rank
of the array.  This has the semantics of first moving the values into
a tuple and then indexing into the array.  The index represented by
the tuple must be an element of the array's domain or an out-of-bounds
error will occur.

\begin{example}
Given the definition
\begin{chapel}
  var ij = (i,j);
\end{chapel}
the indexing expressions \chpl{A(ij)} and \chpl{A(i,j)} are
equivalent.
\end{example}

\subsubsection{Arithmetic Array Slicing}
\label{Arithmetic_Array_Slicing}
\index{arrays!slice}

An arithmetic array can be sliced by any arithmetic domain that is a
subdomain of the array's defining domain.  If the subdomain
relationship is not met, an out-of-bounds error will occur.  The
result is a subarray whose indices are those of the slicing domain and
whose elements are an alias of the original array's.  If the indices
in the slicing

Arithmetic arrays also support slicing by ranges directly.  If each
dimension is indexed by a range, this is equivalent to slicing the
array by the arithmetic domain defined by those ranges.

For multi-dimensional arithmetic arrays, slicing with a rank change is
supported by substituting integral values within a dimension's range
for an actual range.  The resulting array will have a rank less than
the arithmetic array's rank and equal to the number of ranges that are
passed in to take the slice.

\begin{example}
Given an array
\begin{chapel}
var A: [1..n, 1..n] int;
\end{chapel}
the slice \chpl{A[1..n, 1]} is a one-dimensional array whose elements
are the first column of \chpl{A}.
\end{example}

Array slices may also be expressed using partially unbounded or
completely unbounded ranges.  This is equivalent to slicing the
array's defining domain by the specified ranges to create a subdomain
as described in~\rsec{Arithmetic_Domain_Slicing} and then using that
subdomain to slice the array.

\subsubsection{Formal Arguments of Arithmetic Array Type}
\label{Formal_Arguments_of_Arithmetic_Array_Type}
\index{formal arguments!arithmetic arrays}
\index{arrays!as formal arguments}

Formal arguments of arithmetic array type allow an arithmetic domain
to be specified that does not match the arithmetic domain of the
actual arithmetic array that is passed to the formal argument.  In
this case, the shape (size in each dimension and rank) of the domain
of the actual array must match the shape of the domain of the formal
array.  The indices are translated in the formal array, which is a
reference to the actual array.

\begin{example}
In the code
\begin{chapel}
def foo(X: [1..5] int) { ... }
var A: [1..10 by 2] int;
foo(A);
\end{chapel}
the array \chpl{A} is strided and its elements can be indexed by the
odd integers between one and nine.  In the function \chpl{foo}, the
array \chpl{X} references array \chpl{A} and the same elements can be
indexed by the integers between one and five.
\end{example}

\subsection{Sparse Arrays}
\label{Sparse_Arrays}
\index{arrays!sparse}

Sparse domains are used in Chapel to describe irregular index subsets
and to define sparse arrays.  Sparse arrays are typically used to
represent data aggregates in which a value occurs so frequently that
it would be wasteful to store it explicitly for each occurrence.  This
value is commonly described as the ``zero value'', though we refer to
it as the \emph{implicitly replicated value} or \emph{IRV} since it
may be a value other than zero.

\subsubsection{Sparse Arrays}

An array declared over a sparse domain can be indexed using all of the
indices in the domain's parent domain.  If it is read using an index
that is not part of the sparse domain's index set, the IRV value is
returned.  Otherwise, the array's unique value corresponding to the
index is returned.

Sparse arrays can only be written at locations corresponding to
indices in their domain's index set.  In general, writing to other
locations will result in a runtime error.

By default a sparse array's IRV is defined as the default value for
the array's element type.  The IRV can be set to any value of the
array's element type by assigning to a pseudo-field named ``IRV'' in
the array.  It is an error to assign a value to the IRV by assigning
to an array element whose index is not described by the sparse domain.

\begin{example}
The following code example declares a sparse array, \chpl{SpsA} using
the sparse domain \chpl{SpsD} (For this example, assume that
\chpl{n}$>$1).  Lines 2 assigns two indices to \chpl{SpsD}'s index set
and then lines 3--4 store the values 1.1 and 9.9 to the corresponding
values of \chpl{SpsA}.  The IRV of \chpl{SpsA} will initially be 0.0
since its element type is \chpl{real}.  However, the fifth line sets
the IRV to be the value 5.5, causing \chpl{SpsA} to represent the
value 1.1 in its low corner, 9.9 in its high corner, and 5.5
everywhere else.  The final statement is an error since it attempts to
assign to \chpl{SpsA} at an index not described by its domain,
\chpl{SpsD}.

\begin{chapel}
var SpsA: [SpsD] real;
SpsD = ((1,1), (n,n));
SpsA(1,1) = 1.1;
SpsA(n,n) = 9.9;
SpsA.IRV = 5.5;
SpsA(1,n) = 0.0;  // ERROR!
\end{chapel}
\end{example}



\subsection{Associative Arrays}
\label{Associative_Arrays}
\index{arrays!associative}

Arrays declared over associative domains are dictionaries mapping from
values to variables.

\subsection{Association of Arrays to Domains}
\label{Association_of_Arrays_to_Domains}
\index{domains!association to arrays}
\index{arrays!association to domains}

When an array is declared, it is linked during execution to the domain
over which it was declared.  This linkage is constant and cannot be
changed.

When indices are added or removed from a domain, the change impacts
the arrays declared over this particular domain.  In the case of
adding an index, an element is added to the array and initialized to
the default value associated with the element type.  In the case of
removing an index, the element in the array is removed.

When a domain is reassigned a new value, the array is also impacted.
Values that could be indexed by both the old domain and the new domain
are preserved in the array.  Values that could only be indexed by the
old domain are lost.  Values that can only be indexed by the new
domain have elements added to the new array and initialized to the
default value associated with their type.

For performance reasons, there is an expectation that a method will be
added to domains to allow non-preserving assignment, \emph{i.e.}, all
values in the arrays associated with the assigned domain will be lost.

An array's domain can only be modified directly, via the domain's name
or an alias created by passing it to a function via blank intent.  In
particular, the domain may not be modified via the array's
\chpl{.domain} method, nor by using the domain query syntax on a
function's formal array argument
(\rsec{Formal_Arguments_of_Generic_Array_Types}).

\begin{rationale}
When multiple arrays are declared using a single domain, modifying the
domain affects all of the arrays.  Allowing an array's domain to be
queried and then modified suggests that the change should only affect
that array.  By requiring the domain to be modified directly, the user
is encouraged to think in terms of the domain distinctly from a
particular array.

In addition, this choice has the beneficial effect that arrays
declared via an anonymous domain have a constant domain.  Constant
domains are considered a common case and have potential compilation
benefits such as eliminating bounds checks.  Therefore making this
convenient syntax support a common, optimizable case seems prudent.
\end{rationale}

\subsection{Predefined Functions and Methods on Arrays}
\index{arrays!predefined functions}

There is an expectation that this list of predefined functions and
methods will grow.

\begin{protohead}
def $Array$.eltType type
\end{protohead}
\begin{protobody}
Returns the element type of the array.
\end{protobody}

\begin{protohead}
def $Array$.rank param
\end{protohead}
\begin{protobody}
Returns the rank of the array.
\end{protobody}

\begin{protohead}
def $Array$.domain: this.domain
\end{protohead}
\begin{protobody}
Returns the domain of the given array.  This domain is constant,
implying that the domain cannot be resized by assigning to its domain
field, only by modifying the domain directly.
\end{protobody}

\begin{protohead}
def $Array$.numElements: this.domain.dim_type
\end{protohead}
\begin{protobody}
Returns the number of elements in the array.
\end{protobody}

\begin{protohead}
def reshape(A: $Array$, D: $Domain$): $Array$
\end{protohead}
\begin{protobody}
Returns a copy of the array containing the same values but in the
shape of the new domain.  The number of indices in the domain must
equal the number of elements in the array.  The elements of the array
are copied into the new array using the default iteration orders over
both arrays.
\end{protobody}
