\sekshun{Modules}
\label{Modules}
\index{modules}

Chapel supports modules to manage namespaces.  A program consists of
one or more modules.  Every symbol, including variables, functions,
and types, is associated with some module.

Module definitions are described in~\rsec{Module_Definitions}.  The
relation between files and modules is described
in~\rsec{Implicit_Modules}.  Nested modules are described
in~\rsec{Nested_Modules}.  The visibility of a module's symbols by
users of the module is described in~\rsec{Visibility_Of_Symbols}.  The execution
of a program and module initialization/deinitialization are described
in~\rsec{Program_Execution}.

\section{Module Definitions}
\label{Module_Definitions}
\index{module@\chpl{module}}
\index{modules!definitions}

A module is declared with the following syntax:
\begin{syntax}
module-declaration-statement:
  privacy-specifier[OPT] prototype-specifier[OPT] `module' module-identifier block-statement

privacy-specifier:
  `private'
  `public'

prototype-specifier:
  `prototype'

module-identifier:
  identifier
\end{syntax}

A module's name is specified after the \chpl{module} keyword.
The \sntx{block-statement} opens the module's scope.  Symbols defined
in this block statement are defined in the module's scope and are
called \emph{module-scope symbols}.  The visibility of a module is
defined by its \sntx{privacy-specifier}~(\rsec{Visibility_Of_A_Module}).

Module declaration statements are only legal as file-scope or
module-scope statements.  For example, module declaration statements
may not occur within block statements, functions, classes, or records.

\index{modules!top-level}
Any module declaration that is not contained within another module
creates a \emph{top-level module}.  Module declarations within other
modules create nested modules~(\rsec{Nested_Modules}).

\section{Prototype Modules}
\label{Prototype_Modules}
\index{modules!prototype}

Modules that are declared with the \chpl{prototype} keyword use relaxed
rules for error handling and \chpl{nil} checking. These relaxed rules are
appropriate for programs in the early stages of development but are not
appropriate for libraries. In particular, within a \chpl{prototype} module:

\begin{itemize}

  \item errors that are not handled will terminate the program
        (see~\rsec{Errors_Prototype_Mode})

  \item methods on nilable class instances can be called - these will
        implicitly convert to non-nilable class instances
        (see~\rsec{Methods_On_Nilable_In_Prototype_Modules})

\end{itemize}

\section{Files and Implicit Modules}
\label{Implicit_Modules}
\index{modules!and files}

Multiple modules can be defined within the same file and need not bear any
relation to the file in terms of their names.

\begin{chapelexample}{two-modules.chpl}
The following file contains two explicitly named modules, MX and MY.
\begin{chapel}
module MX {
  var x: string = "Module MX";
  proc printX() {
    writeln(x);
  }
}

module MY {
  var y: string = "Module MY";
  proc printY() {
    writeln(y);
  }
}
\end{chapel}
\begin{chapelpost}
MX.printX();
MY.printY();
\end{chapelpost}
\begin{chapeloutput}
Module MX
Module MY
\end{chapeloutput}
Module MX defines module-scope symbols x and printX, while MY
defines module-scope symbols y and printY.
\end{chapelexample}

For any file that contains file-scope statements other than module
declarations, the file itself is treated as a module declaration.
In this case,
\index{implicit modules}
\index{modules!implicit}
the module is implicit. Implicit modules are always \chpl{prototype}
modules. An implicit module takes its name from the base filename.  In
particular, the module name is defined as the remaining string after
removing the \chpl{.chpl} suffix and any path specification from the
specified filename.  If the resulting name is not a legal Chapel
identifier, it cannot be referenced in a use statement.

\begin{chapelexample}{implicit.chpl}
The following file, named implicit.chpl, defines an implicitly named
module called implicit.
\begin{chapel}
var x: int = 0;
var y: int = 1;

proc printX() {
  writeln(x);
}
proc printY() {
  writeln(y);
}
\end{chapel}
\begin{chapelpost}
printX();
printY();
\end{chapelpost}
\begin{chapeloutput}
0
1
\end{chapeloutput}
Module implicit defines the module-scope symbols x, y, printX, and
printY.
\end{chapelexample}


\section{Nested Modules}
\label{Nested_Modules}
\index{modules!nested}
\index{modules!sub-modules}

A \emph{nested module} (or \emph{sub-module}) is a module that is
defined within another module, known as the outer, or parent, module.
A nested module automatically has access to all of the symbols in its
outer module.  However, an outer module needs to explicitly name or
use a nested module in order to access its symbols.

A nested module can be used without using the outer module by
explicitly naming the outer module in the use statement.
\begin{chapelexample}{nested-use.chpl}
The code
\begin{chapelpre}
module libsci {
  writeln("Initializing libsci");
  module blas {
    writeln("\\tInitializing blas");
  }
}
module testmain { // used to avoid warnings
}
\end{chapelpre}
\begin{chapel}
use libsci.blas;
\end{chapel}
\begin{chapeloutput}
Initializing libsci
	Initializing blas
\end{chapeloutput}
uses a module named \chpl{blas} that is nested inside a module
named \chpl{libsci}.
\end{chapelexample}

Files with both module declarations and file-scope statements result in
nested modules.

\begin{chapelexample}{nested.chpl}
The following file, named nested.chpl, defines an
implicitly named module called nested, with nested modules
MX and MY.
\begin{chapel}
module MX {
  var x: int = 0;
}

module MY {
  var y: int = 0;
}

use MX, MY;

proc printX() {
  writeln(x);
}

proc printY() {
  writeln(y);
}
\end{chapel}
\begin{chapelpost}
printX();
printY();
\end{chapelpost}
\begin{chapeloutput}
0
0
\end{chapeloutput}
\end{chapelexample}


\section{Access of Module Contents}
\label{Access_Of_Module_Contents}
\index{modules!access}

A module's contents can be accessed by code outside of that module
depending on the visibility of the module
itself~(\rsec{Visibility_Of_A_Module}) and the visibility of each
individual symbol~(\rsec{Visibility_Of_Symbols}).  This can be done
via the use statement~(\rsec{Using_Modules}) or qualified
naming~(\rsec{Explicit_Naming}).

\subsection{Visibility Of A Module}
\label{Visibility_Of_A_Module}
\index{modules!access}

A top-level module is available for use~(\rsec{Using_Modules}) anywhere. The
visibility of a nested module is subject to the rules
of~\rsec{Visibility_Of_Symbols}, where the nested module is considered
a "module-scope symbol" of its outer module.

\subsection{Visibility Of A Module's Symbols}
\label{Visibility_Of_Symbols}
\index{modules!access}

A symbol defined at module scope is \emph{visible} from outside the
module when the \sntx{privacy-specifier} of its definition
is \chpl{public} or is omitted (i.e. by default). When a module-scope
symbol is declared
\chpl{private}, it is not visible outside of that module. A
symbol's visibility inside its module is controlled by normal lexical
scoping and is not affected by its \sntx{privacy-specifier}.  When a
module's symbol is visible~(\rsec{Visibility_Of_A_Module}), the
visible symbols it contains are accessible via the use
statement~(\rsec{Using_Modules}) or qualified
naming~(\rsec{Explicit_Naming}).

\subsection{Using Modules}
\label{Using_Modules}
\index{modules!using}

The \chpl{use} statement provides the primary means of accessing a
module's symbols from outside of the module.  Use statements make both
the module's name and its public symbols available for reference
within a given scope.  For top-level modules, a use statement is
required before referring to the module's name or the symbols it
contains within a given lexical scope.

Use statements can also restrict or rename the set of module symbols
that are available within the scope.  For further information about
use statements, see~\rsec{The_Use_Statement}.


\subsection{Qualified Naming of Module Symbols}
\label{Explicit_Naming}
\index{modules!qualified naming}

When a module's symbol is visible---via a use statement, or lexically
for nested modules---its public symbols can be referred to via
qualified naming with the following syntax:
\begin{syntax}
module-access-expression:
  module-identifier-list . identifier

module-identifier-list:
  module-identifier
  module-identifier . module-identifier-list

\end{syntax}
This allows two symbols that have the same name to be distinguished
based on the name of their module.  Using qualified naming in a
function call restricts the set of candidate functions to those in the
specified module.

If code refers to symbols that are defined by multiple modules, the
compiler will issue an error.  Qualified naming can be used to
disambiguate the symbols in this case.

\begin{chapelexample}{ambiguity.chpl}
In the following example,
\begin{chapel}
module M1 {
  var x: int = 1;
  var y: int = -1;
  proc printX() {
    writeln("M1's x is: ", x);
  }
  proc printY() {
    writeln("M1's y is: ", y);
  }
}
 
module M2 {
  use M3;
  use M1;

  var x: int = 2;

  proc printX() {
    writeln("M2's x is: ", x);
  }

  proc main() {
    M1.x = 4;
    M1.printX();
    writeln(x);
    printX(); // This is not ambiguous
    printY(); // ERROR: This is ambiguous
  }
}

module M3 {
  var x: int = 3;
  var y: int = -3;
  proc printY() {
    writeln("M3's y is: ", y);
  }
}
\end{chapel}
\begin{chapeloutput}
ambiguity.chpl:22: In function 'main':
ambiguity.chpl:27: error: ambiguous call 'printY()'
ambiguity.chpl:34: note: candidates are: printY()
ambiguity.chpl:7: note:                 printY()
\end{chapeloutput}
The call to printX() is not ambiguous because M2's definition shadows
that of M1.  On the other hand, the call to printY() is ambiguous
because it is defined in both M1 and M3.  This will result in a
compiler error.  The call could be qualified via M1.printY() or M3.printY()
to resolve this ambiguity.
\end{chapelexample}

\subsection{Module Initialization}
\label{Module_Initialization}
\index{modules!initialization}
\index{initialization!modules}

Module initialization occurs at program start-up.  All module-scope
statements within a module other than function and type declarations are
executed during module initialization. Modules that are not referred to,
including both top-level modules and sub-modules, will not be initialized.

\begin{chapelexample}{init.chpl}
In the code,
\begin{chapelpre}
proc foo() {
    return 1;
}
\end{chapelpre}
\begin{chapel}
var x = foo();       // executed at module initialization
writeln("Hi!");      // executed at module initialization
proc sayGoodbye {
  writeln("Bye!");   // not executed at module initialization
}
\end{chapel}
\begin{chapeloutput}
Hi!
\end{chapeloutput}
The function foo() will be invoked and its result assigned to x.  Then
``Hi!'' will be printed.
\end{chapelexample}

Module initialization order is discussed
in~\rsec{Module_Initialization_Order}.


\subsection{Module Deinitialization}
\label{Module_Deinitialization}
\index{modules!deinitialization}
\index{deinitialization!modules}

Module deinitialization occurs at program tear-down.
During module deinitialization:

\begin{itemize}

\item If the module contains a deinitializer, which is a module-scope function
named \chpl{deinit()}, it is executed first.

\item If the module declares global variables, they are deinitialized
in the reverse declaration order.

\end{itemize}

% TODO: add examples. For example, adapt this test:
%   test/modules/vass/deinit-order-modules.chpl

Module deinitialization order is discussed
in~\rsec{Module_Deinitialization_Order}.


\section{Program Execution}
\label{Program_Execution}
\index{program execution}
\index{program initialization}

Chapel programs start by initializing all modules and then executing
the main function~(\rsec{The_main_Function}).

\subsection{The {\em main} Function}
\label{The_main_Function}

\index{main@\chpl{main}}
\index{functions!main@\chpl{main}}
The main function must be called \chpl{main} and must have zero
arguments.  It can be specified with or without parentheses.  In any
Chapel program, there is a single main function that defines the
program's entry point.  If a program defines multiple potential entry
points, the implementation may provide a compiler flag that
disambiguates between main functions in multiple modules.

\begin{craychapel}
In the Cray Chapel compiler implementation, the \emph{--
--main-module} flag can be used to specify the module from which the
main function definition will be used.
\end{craychapel}

\begin{chapelexample}{main-module.chpl}
Because it defines two \chpl{main} functions, the following code will yield an
error unless a main module is specified on the command line.
\begin{chapel}
module M1 {
  const x = 1;
  proc main() {
    writeln("M", x, "'s main");
  }
}
 
module M2 {
  use M1;

  const x = 2;
  proc main() {
    M1.main();
    writeln("M", x, "'s main");
  }
}
\end{chapel}
\begin{chapelcompopts}
--main-module M1 \# main\_module.M1.good
--main-module M2 \# main\_module.M2.good
\end{chapelcompopts}
If M1 is specified as the main module, the program will output:
\begin{chapelprintoutput}{main_module.M1.good}
M1's main
\end{chapelprintoutput}
If M2 is specified as the main module the program will output:
\begin{chapelprintoutput}{main_module.M2.good}
M1's main
M2's main
\end{chapelprintoutput}
Notice that main is treated like just another function if it is not in
the main module and can be called as such.
\end{chapelexample}

\index{exploratory programming}

%subsubsection{Programs with a Single Module}
%% \label{Programs_with_a_Single_Module}

To aid in exploratory programming, a default main function is
created if the program does not contain a user-defined main function.  The
default main function is equivalent to
\begin{chapel}
proc main() {}
\end{chapel}

\begin{chapelexample}{no-main.chpl}
The code
\begin{chapel}
writeln("hello, world");
\end{chapel}
\begin{chapeloutput}
hello, world
\end{chapeloutput}
is a legal and complete Chapel program.  The startup code for a Chapel program
first calls the module initialization code for the main module and then
calls \chpl{main()}.  This program's initialization function is the file-scope
writeln() statement.  The module declaration is taken to be the entire file,
as described in~\rsec{Implicit_Modules}.
\end{chapelexample}


\subsection{Module Initialization Order}
\label{Module_Initialization_Order}
\index{modules!initialization order}

Module initialization is performed using the following algorithm.

Starting from the module that defines the main function, the modules named in
its use statements are visited depth-first and initialized in post-order.  If a
use statement names a module that has already been visited, it is not visited a
second time.  Thus, infinite recursion is avoided.

Modules used by a given module are visited in the order in which
they appear in the program text.  For nested modules, the
parent module and its uses are initialized before the nested module and its uses.

\begin{chapelexample}{init-order.chpl}
The code
\begin{chapel}
module M1 {
  use M2.M3;
  use M2;
  writeln("In M1's initializer");
  proc main() {
    writeln("In main");
  }
}

module M2 {
  use M4;
  writeln("In M2's initializer");
  module M3 {
    writeln("In M3's initializer");
  }
}

module M4 {
  writeln("In M4's initializer");
}
\end{chapel}
prints the following
\begin{chapelprintoutput}{}
In M4's initializer
In M2's initializer
In M3's initializer
In M1's initializer
In main
\end{chapelprintoutput}
M1, the main module, uses M2.M3 and then M2, thus M2.M3 must be
initialized.  Because M2.M3 is a nested module, M4 (which is used by
M2) must be initialized first.  M2 itself is initialized, followed by
M2.M3.  Finally M1 is initialized, and the main function is run.
\end{chapelexample}


\subsection{Module Deinitialization Order}
\label{Module_Deinitialization_Order}
\index{modules!deinitialization order}

Module deinitialization is performed in the reverse order of
module initialization, as specified in
\rsec{Module_Initialization_Order}.

% TODO: add examples. For example, adapt this test:
%   test/modules/vass/deinit-order-modules.chpl
